---
title: "Course 6 - Project - Part 2"
author: "Bruno Assis"
date: "19/03/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Impact of Vitamin C in the Tooth Growth of Guinea Pigs

## Overview

Now in the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package.

1. Load the ToothGrowth data and perform some basic exploratory data analyses
2. Provide a basic summary of the data.
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
4. State your conclusions and the assumptions needed for your conclusions.

## Loading ToothGrowth Data

```{r}
data("ToothGrowth")
```

## Exploratory Data Analyses

```{r}
tooth.df <- data.frame(ToothGrowth)
dim(tooth.df)
head(tooth.df)
```

Let's convert the `supplement` and `dose` attributes to factors:
```{r}
tooth.df$supp <- factor(tooth.df$supp)
tooth.df$dose <- factor(tooth.df$dose)
```

Let's plot the data:
```{r}
ggplot(tooth.df, aes(x=supp, y=len)) +
  geom_boxplot(aes(fill=supp))
```

### Basic Summary of Data
## Conclusions